Double Standards in Judging Collective Action\thanks{}

### 2022-08-16 Documents/manuscript.tex
first complete draft

%\date{June 15, 2021}

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \noindent Collective action is a powerful force driving social
change but often sparks contention about what actions are acceptable
means to effect social change. In four studies (total \emph{N} = 2,246),
we investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In two
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police in the United States. We
found evidence for ideology-based, but not for identity-based, double
standards: Participants judged the same protest actions to be more
acceptable when the protesters' cause aligned with their own
(system-challenging) ideological position---but participants showed no
ingroup bias when judging collective action by ingroup and outgroup
members. Our findings have theoretical and practical implications for
understanding the often divided response to prominent social movements.
    \end{abstract}
  \end{@twocolumnfalse}
  % \renewcommand{\abstractname}{Statement of Relevance}
  % \begin{@twocolumnfalse}
  %   \maketitle
  %   \begin{abstract}
  %   \noindent 
  %   \end{abstract}
  % \end{@twocolumnfalse}
]

@@

The second purpose of our research is to test for double standards in
judging collective action---that is, whether people judge the same
protest actions as more acceptable if the protesters' are ingroup
members or the protesters' cause aligns with the ingroup's interests
(Hypothesis 1a, 1b) or if the protesters' cause aligns with their own
system-justifying or system-challenging ideological orientation
(Hypotheses 2a, 2b). In Experiment 1, we use the instrument developed in
Studies 1 and 2 to test for double standards in how acceptable
participants consider collective action for workers' rights in the
United Kingdom. By varying both the participants' and the protesters'
group memberships (working/middle class), Experiment 1 compares the
relative evidence for identity-based (Hypotheses 1a, 1b) and
ideology-based double standards (Hypothesis 2a). Extending Experiment 1,
Experiment 2 considers both system-challenging and system-defending
collective action in the context of the 2020 Black Lives Matter protests
in the United States. By varying the protesters' cause (for/against
defunding the police) as well as the participants' and the protesters'
group memberships (Black/White), Experiment 2 compares the relative
evidence identity-based (Hypotheses 1a, 1b) and ideology-based double
standards (Hypothesis 2a, 2b). Together, the preregistered experiments
provide a comprehensive test of the hypothesized double standards in
judging collective action.

@@

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that, in line with Hypothesis 1a and
1b, participants would judge the same protest actions as more acceptable
if the protesters were ingroup members protesting for the ingroup's
rights than if they were outgroup members protesting for the outgroup's
rights. By comparing reactions to protesters with working-class or
professional jobs, we tested how observers judge collective action by
lower-status (i.e.~working-class) and higher-status
(professional/middle-class) group members. Assuming that collective
action by lower-status group members is more threatening to the unequal
status than collective action by higher-status group members, we tested
the preregistered hypothesis that, in line with Hypothesis 2a,
participants would judge the same protest actions as more acceptable if
the protesters were from a higher-status group than if they were from a
lower-status group. In non-preregistered analyses, we further explored
whether, in line with Hypothesis 2b, participants would judge the same
protest actions as more or less acceptable depending on their own
ideological positions. In this way, Experiment 1 provided a first test
of the hypothesized identity- and ideology-based double standards in
judging collective action.

@@

We preregistered the sample size as well as all hypotheses,
inclusion/exclusion criteria, measures, and manipulations.\footnote{\url{https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b}}
We made all materials, data, and analysis scripts available
online.\footnote{\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}}
We followed sample size recommendations for item response theory models
(DeMars, 2010, p. 36), planning to recruit 500 participants.

@@

Figure \ref{fig:f3} (A) shows estimates for each combination of the
protesters' and the participants' group membership. Overall, we found
that participants' responses depended on both the protesters' and the
participants' group status---but not in the directions predicted by our
hypotheses. Contradicting Hypothesis 1, participants did not consider
protest actions performed by their ingroup to be more acceptable, on
average, than the same actions performed by the relevant outgroup
(Cohen's \(d = 0.00, [-0.20, 0.20]\)). Instead, we found that
participants from higher-status backgrounds considered protest actions
performed by both lower-status (\(d = 0.34, [0.07, 0.59\){]}) and
higher-status (\(d = 0.33, [0.03, 0.63]\)) protesters to be, on average,
more acceptable than participants from lower-status backgrounds.
Contradicting Hypothesis 2, participants considered protest actions
performed by higher-status protesters to be less acceptable, on average,
than the same actions performed by lower-status protesters
(\(d = -0.20, [-0.40, -0.00]\)).

@@

\ldots

@@

In Experiment 2, we investigated potential double standards in judging
collective action for or against defunding the police in the United
States. That is, we tested whether Black and White Americans applied
different standards when judging collective action taken by either Black
or White protesters to protest either for or against police divestment
as a possible solution to racist police violence.

\begin{itemize}
\tightlist
\item
  Explain why defunding the police (public support for BLM, so needed to
  be more controversial)
\end{itemize}

@@

We preregistered the sample size as well as all hypotheses,
inclusion/exclusion criteria, measures, and manipulations.\footnote{\url{https://osf.io/skxjt/?view_only=d8ce44a700884b5ab36b64ef08f833a1}}
We made all materials, data, and analysis scripts available
online.\footnote{\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}}
As reported in the preregistration, we ran simulations, using data from
Experiment 1, to determine that a sample size of \(N = 1,600\)
(\(n = 200\) per condition) was sufficient to detect even small
differences between conditions (\(0.09 < \text{Cohen's}~d < 0.16\)).

@@

As expected, Black participants (\(M = 6.18, \textit{SD} = 1.31\))
reported being more outraged about recent incidents of police violence
than White participants (\(M = 5.56, \textit{SD} = 1.61\); Cohen's
\(d = 0.42\)). Black participants strongly identified with their ingroup
(\(M = 6.46, \textit{SD} = 1.09\)) and, across experimental conditions,
identified more with Black protesters (\(M = 4.66, \textit{SD} = 1.99\))
than with White protesters (\(M = 4.03, \textit{SD} = 2.16\);
\(d = 0.30\)). In contrast, White participants identified less strongly
with their ingroup (\(M = 4.85, \textit{SD} = 1.53\); \(d = 1.21\)) and
did not identify more with White protesters
(\(M = 3.58, \textit{SD} = 2.01\)) than with Black protesters
(\(M = 3.81, \textit{SD} = 1.77\); \(d = -0.12\)). Both White
(\(d = 0.67\)) and Black (\(d = 0.53\)) participants tended to identify
more with protesters protesting \emph{for} defunding the police. On
average, participants tended to describe their political orientation as
moderately liberal (\(M = 2.92, \textit{SD} = 1.73\)) but, as expected,
were divided in their support for defunding the police
(\(M = 3.37, \textit{SD} = 1.42\)).

@@

Model 1 (M1) estimated varying intercepts for the eight conditions to
test whether, in line with Hypotheses 1a, 1b, or 2a, participants'
responses depended on their own group membership, the protesters' group
membership, or the protesters' cause. Contradicting Hypothesis 1a,
participants did not consider protest actions performed by their ingroup
to be more acceptable, on average, than the same actions performed by
the relevant outgroup (\(d = 0.01, [-0.04, 0.06]\)). Contradicting
Hypothesis 1b, participants did not consider protest actions for a cause
that was nominally aligned with their ingroup's interests to be more
acceptable, on average, than protest actions for a cause not aligned
with their ingroup's interests (\(d = -0.01, [-0.06, 0.04]\)).
Contradicting Hypothesis 2a, participants did not consider protest
actions for a system-defending cause to be more acceptable, on average,
than protest actions for a system-challenging cause
(\(d = -0.14, [-0.20, -0.09]\)). Instead, we found that, on average,
White participants considered all protest actions to be more acceptable
than Black participants (\(d = 0.09, [0.04, 0.14]\)) and both Black
(\(d = 0.07, [0.03, 0.10]\)) and White (\(d = 0.08, [0.04, 0.11]\))
participants considered the same actions to be more acceptable when
protests were for, rather than against, defunding the police. As in
Experiment 1, we thus found that participants' responses depended on
both the participants' and the protesters' group memberships---but not
in the directions predicted by our hypotheses.

Model 2 (M2) extended Model 1 by estimating participants' responses as a
function of their \emph{z}-standardized endorsement of system-justifying
beliefs. As preregistered, we modeled this relationship with two fixed
effects, one estimating the effect of system-justifying beliefs on
judgements about system-defending protest actions and one estimating
their effect on judgements about system-challenging protest actions, and
one varying effect estimating its variance across conditions. Supporting
Hypothesis 2b, participants who \emph{rejected} system-justifying
beliefs were \emph{more} likely to consider system-challenging protest
actions (for defunding the police) to be acceptable means of protest
(\(\beta_{xy} = 0.38, [0.16, 0.58]\)). We did not, however, find
evidence for the ideological symmetry implied by Hypothesis 2b since
participants who \emph{endorsed} system-justifying beliefs were
\emph{not} more likely to consider system-defending protest actions
(against defunding the police) to be acceptable means of protest
(\(\beta_{xy} = -0.18, [-0.40, 0.02]\)).

Figure \ref{fig:f5} shows the estimated pattern of condition-wise
differences underlying those fixed effects. Participants who endorsed
system-justifying beliefs tended to consider system-challenging and
system-defending protest actions to be equally unacceptable. In
contrast, participants who rejected system-justifying beliefs evinced
ideology-based double standards: In line with Hypothesis 2b, they
considered the same protest actions to be more acceptable when the
protesters challenged the system or, to a lesser extent, when the
protesters were from the disadvantaged group.

Model 3 (M3) extended Model 1 by estimating participants' responses as a
function of their self-reported support for the cause of the protest. We
recoded participants' responses to create a predictor variable that
encoded support for defunding the police when protesters supported
defunding the police and opposition to defunding the police when
protesters opposed defunding the police. As preregistered, we modeled
this relationship as a monotonic effect (Bürkner \& Charpentier, 2020)
that estimated the average change in the outcome variable across
predictor categories as well as how much of this change occurred between
each of the four pairs of adjacent predictor categories.\footnote{Our
  model assigned a Dirichlet prior, \(\alpha = {1, 1, 1, 1}\), to the
  proportions of the overall change that was expected to occur between
  each of the four pairs of predictor categories.} Contradicting
Hypothesis 3, participants who supported the protesters' cause did not,
on average, consider the same protest actions to be more acceptable than
participants who opposed the protesters' cause
(\(\beta_{y} = 0.14, [-0.12, 0.40]\)). Ergo, our results did not support
the alternative hypothesis that ideology-based double standards can be
reduced to support for, or opposition to, the cause of a protest.

@@

We reran Model 2 with political orientation, instead of system
justification, as an alternative predictor variable. Our results
mirrored results from the preregistered analyses: More liberal
participants were more likely to consider protest actions for defunding
the police to be acceptable (\(\beta_{xy} = 0.41, [0.17, 0.62]\)) but
more conservative participants were not more likely to consider protest
actions against defunding the police to be acceptable
(\(\beta_{xy} = -0.16, [-0.41, 0.04]\)). As Figure 5 shows, conservative
participants tended to consider all protest actions to be equally
unacceptable while liberal participants considered the same protest
actions to be more acceptable when protesters rallied around a
progressive cause. Our non-preregistered analyses thus replicated the
ideology-based double standards from the preregistered analyses with a
different operationalization of political ideology.

@@

\ldots

@@

\begin{itemize}
\tightlist
\item
  Summary: No evidence for identity-based double standards, but some
  evidence for ideology-based double standards
\item
  Reject the alternative hypothesis that there is consensus about what
  is an acceptable means of protest.
\item
  Relate our findings to Verkuyten et al. (2022)
\item
  Two plausible interpretations: (1) conservatives oppose controversial
  protest (no matter the end) \emph{or} (2) the end justifies the mean
\item
  Re (1) it's plausible but what about recent events? (see public
  polling from grant)
\item
  Re (2) why is there no asymmetry?
\item
  Outline the-end-justifies-the-mean idea, highlighting that just
  supporting a cause is not enough, it needs moralization
\item
  Discuss why high-status group are more accepting of controversial
  protest actions
\item
  Discuss limitations (two cultures, two issues, liberal samples;
  system-defending vs.~system-supporting actions; what about other
  conservative causes?; ingroup bias might need zero-sum situation to
  succeed)
\item
  Invite researchers to use our paradigm
\end{itemize}

@@

@@

>>>>

\date{July 15, 2022}

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \noindent Collective action is a powerful force driving social
change but often sparks contention about what actions are acceptable
means to effect social change. In four studies (total \emph{N} = 2,246),
we investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In two
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police in the United States. We
found evidence for ideology-based, but not for identity-based, double
standards: Participants judged the same protest actions to be more
acceptable when the protesters' cause aligned with their own
(system-challenging) ideological position---but showed no ingroup bias
when judging collective action by ingroup and outgroup members. Our
findings have theoretical and practical implications for understanding
the often divided response to prominent social movements.
    \end{abstract}
  \end{@twocolumnfalse}
  % \renewcommand{\abstractname}{Statement of Relevance}
  % \begin{@twocolumnfalse}
  %   \maketitle
  %   \begin{abstract}
  %   \noindent 
  %   \end{abstract}
  % \end{@twocolumnfalse}
]

@@

The second purpose of our research is to test for double standards in
judging collective action---that is, whether people judge the same
protest actions as more acceptable if the protesters' are ingroup
members or the protesters' cause aligns with the ingroup's interests
(\emph{Hypothesis 1a}, \emph{1b}) or if the protesters' cause aligns
with their own system-justifying or system-challenging motives
(\emph{Hypotheses 2a}, \emph{2b}). In Experiment 1, we use the
instrument developed in Studies 1 and 2 to test for double standards in
how acceptable participants consider collective action for workers'
rights in the United Kingdom. By varying both the participants' and the
protesters' group memberships (working/middle class), Experiment 1
compares the relative evidence for identity-based (\emph{Hypothesis 1a},
\emph{1b}) and ideology-based double standards (\emph{Hypothesis 2a}).
Extending Experiment 1, Experiment 2 considers both system-challenging
and system-defending collective action in the context of the 2020 Black
Lives Matter protests in the United States. By varying the protesters'
cause (for/against defunding the police) as well as the participants'
and the protesters' group memberships (Black/White), Experiment 2
compares the relative evidence identity-based (\emph{Hypothesis 1a},
\emph{1b}) and ideology-based double standards (\emph{Hypothesis 2a},
\emph{2b}). Together, the preregistered experiments provide a
comprehensive test of the hypothesized double standards in judging
collective action.

@@

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that, in line with \emph{Hypothesis
1a} and \emph{1b}, participants would judge the same protest actions as
more acceptable if the protesters were ingroup members protesting for
the ingroup's rights than if they were outgroup members protesting for
the outgroup's rights. By comparing reactions to working-class and
middle-class protesters, we tested how observers judge collective action
by lower-status (i.e.~working-class) and higher-status (middle-class)
group members. Assuming that collective action by lower-status group
members is seen as more threatening to the unequal status quo than
collective action by higher-status group members and that people are, to
same extent, motivated to justify the system, we tested the
preregistered hypothesis that, in line with \emph{Hypothesis 2a},
participants would judge the same protest actions as more acceptable if
the protesters were from a higher-status group than if they were from a
lower-status group. In non-preregistered analyses, we further explored
whether, in line with \emph{Hypothesis 2b}, participants would judge the
same protest actions as more or less acceptable depending on their own
ideological positions. In this way, Experiment 1 provided a first test
of the hypothesized identity- and ideology-based double standards in
judging collective action.

@@

We preregistered the sample size as well as all hypotheses,
inclusion/exclusion criteria, measures, and manipulations
(\url{https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b}).
We made all materials, data, and analysis scripts available online
(\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}).
We followed sample size recommendations for item response theory models
(DeMars, 2010, p. 36), planning to recruit 500 participants.

@@

Figure \ref{fig:f3} (A) shows estimates for each combination of the
protesters' and the participants' group membership. Overall, we found
that participants' responses depended on both the protesters' and the
participants' group status---but not in the directions predicted by our
hypotheses. Contradicting our first hypothesis, participants did not
consider protest actions performed by their ingroup to be more
acceptable, on average, than the same actions performed by the relevant
outgroup (Cohen's \(d = 0.00, [-0.20, 0.20]\)). Instead, we found that
participants from higher-status backgrounds considered protest actions
performed by both lower-status (\(d = 0.34, [0.07, 0.59\){]}) and
higher-status (\(d = 0.33, [0.03, 0.63]\)) protesters to be, on average,
more acceptable than participants from lower-status backgrounds.
Contradicting our second hypothesis, participants considered protest
actions performed by higher-status protesters to be less acceptable, on
average, than the same actions performed by lower-status protesters
(\(d = -0.20, [-0.40, -0.00]\)).

@@

Experiment 1 provided a first test of the hypothesized identity- and
ideology-based double standards in judging collective action. That is,
we tested whether, in line with \emph{Hypothesis 1a} and \emph{1b},
participants would judge the same protest actions as more acceptable if
the protesters were ingroup members protesting for the ingroup's rights
or whether, in line with \emph{Hypothesis 2a}, participants would judge
the same protest actions as more acceptable if the protesters were from
a higher-status group. We found that participants' responses depended on
both the participants' and the protesters' group memberships---but not
in the hypothesized directions. Instead, we found that both lower-status
and higher-status participants tended to find the same protest actions
more acceptable when lower-status protesters protested a bill
threatening their rights. In non-preregistered analyses, we further
found that people who were right-wing and endorsed system-justifying
beliefs tended to find all protest actions to be less acceptable.

Experiment 1 did not, however, provide a conclusive test of our
hypotheses. First, by focusing on fictitious government bills, we might
have chosen a scenario too far removed from current affairs to evoke
strong ingroup bias in participants' judgements of protest actions. That
said, participants' reactions to the experimental manipulation suggested
that it evoked outrage and was understood in ingroup--outgroup terms.
Second, by focusing on collective action to defend workers' rights, we
arguably studied reactions to collective action for a progressive cause.
To test for ideology-based double standards, we relied on the assumption
that collective action by lower-status group members would be seen as
more challenging to the prevailing system than collective action by
higher-status group members. This assumption, however, was contradicted
by our observation that participants who endorsed system-justifying
beliefs tended to find actions by lower-status and higher-status
protesters to be less acceptable. Experiment 2 addressed those
limitations by focusing on a scenario we expected to evoke stronger
responses from participants and by examining reactions to both
system-challenging and system-defending collective action.

@@

In Experiment 2, we investigated potential double standards in judging
collective action for or against defunding the police in the United
States. That is, we tested whether Black and White Americans applied
different standards when judging collective action taken by either Black
or White protesters to protest either for or against police divestment
as a possible solution to racist police violence. We conducted this
study in January 2021, a time when, after a year of unprecedented
protests for racial justice, most Americans could be expected to be
aware of, and have formed an opinion on, the Black Lives Matter
movement. We focused on defunding the police as this position remained
controversial among both Black and White Americans even as they broadly
agreed on the need for police reform.\footnote{In December 2020, only
  34\% of Black Americans and 18\% of White Americans supported
  defunding the police while majorities of Black and White Americans
  supported police reforms such as eliminating qualified immunity (50\%
  and 57\%) or banning chokeholds (79\% and 61\%, YouGov, 2020).}

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that participants would judge the
same protest actions as more acceptable if the protesters were ingroup
members (\emph{Hypothesis 1a}). By assuming that defunding the police
aligns more closely with the interests of Black than White Americans, we
tested the preregistered hypothesis that participants would judge the
same protest actions as more acceptable if the cause of the protest
aligned with their ingroup's interests (\emph{Hypothesis 1b}). By
varying whether the protesters protested for or against defunding the
police, we tested the preregistered hypotheses that participants would
judge the same protest actions as more acceptable if the protest
supported, rather than challenged, the system (\emph{Hypothesis 2a}) and
if they endorsed system-justifying beliefs and the protest supported the
system or if they rejected system-justifying beliefs and the protest
challenged the system (\emph{Hypothesis 2b}). In this way, Experiment 2
provided a complete test of the hypothesized identity- and
ideology-based double standards in judging collective action. In
addition, we tested the simpler, alternative hypothesis that
participants, in general, would judge the same protest actions as more
acceptable if they supported the protesters' cause (\emph{Hypothesis
3}).

@@

We preregistered the sample size as well as all hypotheses,
inclusion/exclusion criteria, measures, and manipulations
(\url{https://osf.io/skxjt/?view_only=d8ce44a700884b5ab36b64ef08f833a1}).
We made all materials, data, and analysis scripts available online
(\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}).
As reported in the preregistration, we ran simulations, using data from
Experiment 1, to determine that a sample size of \(N = 1,600\)
(\(n = 200\) per condition) was sufficient to detect even small
differences between conditions (\(0.09 < \text{Cohen's}~d < 0.16\)).

@@

As expected, Black participants (\(M = 6.18, \textit{SD} = 1.31\))
reported being more outraged about recent incidents of police violence
than White participants (\(M = 5.56, \textit{SD} = 1.61\); Cohen's
\(d = 0.42\)). Black participants strongly identified with their ingroup
(\(M = 6.46, \textit{SD} = 1.09\)) and, across experimental conditions,
identified more with Black protesters (\(M = 4.66, \textit{SD} = 1.99\))
than with White protesters (\(M = 4.03, \textit{SD} = 2.16\);
\(d = 0.30\)). In contrast, White participants identified less strongly
with their ingroup (\(M = 4.85, \textit{SD} = 1.53\); \(d = 1.21\)) and
did not identify more with White protesters
(\(M = 3.58, \textit{SD} = 2.01\)) than with Black protesters
(\(M = 3.81, \textit{SD} = 1.77\); \(d = -0.12\)). Both White
(\(d = 0.67\)) and Black (\(d = 0.53\)) participants tended to identify
more with protesters protesting \emph{for} defunding the police. On
average, participants tended to describe their political orientation as
moderately liberal (\(M = 2.92, \textit{SD} = 1.73\)) but, as expected,
were divided in their support for defunding the police
(\(M = 3.37, \textit{SD} = 1.42\)). That is, 56\% supported defunding
the police, 30\% opposed defunding the police, and 14\% remained
undecided.

@@

Model 1 (M1) estimated varying intercepts for the eight conditions to
test whether, in line with Hypotheses 1a, 1b, or 2a, participants'
responses depended on their own group membership, the protesters' group
membership, or the protesters' cause. Contradicting \emph{Hypothesis
1a}, participants did not consider protest actions performed by their
ingroup to be more acceptable, on average, than the same actions
performed by the relevant outgroup (\(d = 0.01, [-0.04, 0.06]\)).
Contradicting \emph{Hypothesis 1b}, participants did not consider
protest actions for a cause that was nominally aligned with their
ingroup's interests to be more acceptable, on average, than protest
actions for a cause not aligned with their ingroup's interests
(\(d = -0.01, [-0.06, 0.04]\)). Contradicting \emph{Hypothesis 2a},
participants did not consider protest actions for a system-defending
cause to be more acceptable, on average, than protest actions for a
system-challenging cause (\(d = -0.14, [-0.20, -0.09]\)). Instead, we
found that, on average, White participants considered all protest
actions to be more acceptable than Black participants
(\(d = 0.09, [0.04, 0.14]\)) and both Black (\(d = 0.07, [0.03, 0.10]\))
and White (\(d = 0.08, [0.04, 0.11]\)) participants considered the same
actions to be more acceptable when protests were for, rather than
against, defunding the police. As in Experiment 1, we thus found that
participants' responses depended on both the participants' and the
protesters' group memberships---but not in the directions predicted by
our hypotheses.

Model 2 (M2) extended Model 1 by estimating participants' responses as a
function of their \emph{z}-standardized endorsement of system-justifying
beliefs. As preregistered, we modeled this relationship with two fixed
effects, one estimating the effect of system-justifying beliefs on
judgements about system-defending protest actions and one estimating
their effect on judgements about system-challenging protest actions, and
one varying effect estimating its variance across conditions. Supporting
\emph{Hypothesis 2b}, participants who \emph{rejected} system-justifying
beliefs were \emph{more} likely to consider system-challenging protest
actions (for defunding the police) to be acceptable means of protest
(\(\beta_{xy} = 0.38, [0.16, 0.58]\)). We did not, however, find
evidence for the ideological symmetry implied by \emph{Hypothesis 2b}
since participants who \emph{endorsed} system-justifying beliefs were
\emph{not} more likely to consider system-defending protest actions
(against defunding the police) to be acceptable means of protest
(\(\beta_{xy} = -0.18, [-0.40, 0.02]\)).

Figure \ref{fig:f5} shows the estimated pattern of condition-wise
differences underlying those fixed effects. Participants who endorsed
system-justifying beliefs tended to consider system-challenging and
system-defending protest actions to be equally unacceptable. In
contrast, participants who rejected system-justifying beliefs evinced
ideology-based double standards: In line with \emph{Hypothesis 2b}, they
considered the same protest actions to be more acceptable when the
protesters challenged the system or, to a lesser extent, when the
protesters were from the disadvantaged group.

Model 3 (M3) extended Model 1 by estimating participants' responses as a
function of their self-reported support for the cause of the protest. We
recoded participants' responses to create a predictor variable that
encoded support for defunding the police when protesters supported
defunding the police and opposition to defunding the police when
protesters opposed defunding the police. As preregistered, we modeled
this relationship as a monotonic effect (Bürkner \& Charpentier, 2020)
that estimated the average change in the outcome variable across
predictor categories as well as how much of this change occurred between
each of the four pairs of adjacent predictor categories.\footnote{Our
  model assigned a Dirichlet prior, \(\alpha = {1, 1, 1, 1}\), to the
  proportions of the overall change that was expected to occur between
  each of the four pairs of predictor categories.} Contradicting
\emph{Hypothesis 3}, participants who supported the protesters' cause
did not, on average, consider the same protest actions to be more
acceptable than participants who opposed the protesters' cause
(\(\beta_{y} = 0.14, [-0.12, 0.40]\)). Ergo, our results did not support
the alternative hypothesis that ideology-based double standards can be
reduced to support for, or opposition to, the cause of a protest.

@@

We explored whether group identification moderated how the participants'
and the protesters' group memberships affected participants' responses.
To that end, we extended Model 1 by estimating participants' responses
as a function of their \emph{z}-standardized identification with their
racial ingroup. As in Model 2, we modeled this relationship with a fixed
and a varying effect. We found, however, that even participants who
strongly identified with their racial ingroup (+1\emph{SD}) did not
consider protest actions performed by their ingroup to be more
acceptable than the same actions performed by the relevant outgroup
(\(d = 0.02, [-0.04, 0.07]\)) and did not consider protest actions for a
cause that was aligned with their ingroup's interests to be more
acceptable than protest actions for a cause not aligned with their
ingroup's interests (\(d = -0.03, [-0.09, 0.02]\)). Our
non-preregistered analyses thus suggested that group identification did
not moderate group differences in judgements about collective action by
ingroup and outgroup members.

We also explored whether we would find ideology-based double standards
when operationalizing ideology as political orientation instead of as
system justification. To that end, we reran Model 2 with political
orientation as the \emph{z}-standardized predictor variable. Our results
mirrored the preregistered analyses: More liberal participants were more
likely to consider protest actions for defunding the police to be
acceptable (\(\beta_{xy} = 0.41, [0.17, 0.62]\)) but more conservative
participants were not more likely to consider protest actions against
defunding the police to be acceptable
(\(\beta_{xy} = -0.16, [-0.41, 0.04]\)). As Figure 5 shows, conservative
participants tended to consider all protest actions to be equally
unacceptable while liberal participants considered the same protest
actions to be more acceptable when protesters rallied around a
progressive cause. Our non-preregistered analyses thus replicated the
ideology-based double standards from the preregistered analyses with a
different operationalization of ideology.

@@

Experiment 2 provided a complete test of the hypothesized identity- and
ideology-based double standards in judging collective action. As in
Experiment 1, we found that participants' responses depended on both the
participants' and the protesters' group memberships---but not in the
hypothesized directions. We found that, contrary to \emph{Hypothesis
1a}, participants considered protest actions taken by ingroup and
outgroup members to be equally acceptable and that, contrary to
\emph{Hypothesis 1b}, participants did not consider the same protest
actions to be more acceptable when the protesters' cause aligned with
their ingroup's interests. Like Experiment 1, Experiment 2 thus found no
evidence for identity-based double standards---even though it focused on
an issue of direct and current relevance to the participants.

Expanding Experiment 1, Experiment 2 considered reactions to both
system-challenging and system-defending collective action and, in so
doing, provided a stronger test of the hypothesized ideology-based
double standards. Contrary to \emph{Hypothesis 2a} and the underlying
assumption of a universal system-justifying motive, we found that
participants considered the same protest actions to be \emph{more}
acceptable when judging system-challenging collective action (for
defunding the police) than when judging system-defending collective
action. This finding might reflect our liberal-leaning sample. In line
with \emph{Hypothesis 2b}, we found that participants who rejected
system-justifying beliefs considered the same protest action more
acceptable when judging system-challenging collective action (for
defunding the police). We did not, however, find evidence for
ideological symmetry in this relationship as participants who endorsed
system-justifying beliefs considered system-challenging and
system-defending collective action to be equally unacceptable. In
addition, we did not find evidence for the alternative, simpler
\emph{Hypothesis 3} that that participants would judge the same protest
actions to be acceptable when they supported the protesters' cause. In
this way, Experiment 2 provided evidence for ideology-based, but not for
identity-based, double standards in judging collective action.

@@

Our research was motivated by real-world examples that seemed to suggest
that where people draw the line between acceptable and unacceptable
protest actions depends on \emph{who} the protesters are and \emph{what}
they are protesting---in other words, that there are double standards in
judging collective action. Our findings, however, showed that it is not
so simple. Contradicting our hypothesis of identity-based double
standards, participants in two preregistered experiments did not show
ingroup bias when judging controversial protest actions taken by ingroup
and outgroup members to advance their group's interests. Instead,
supporting our hypothesis of ideology-based double standards,
participants judged the same controversial protest actions to be more
acceptable when the protesters' cause aligned with their own
(system-challenging) ideological position. In the remainder of this
General Discussion, we discuss theoretical and practical implications
for understanding the often divided response to collective action and
consider some limiting conditions on our findings.

\hypertarget{implications}{%
\subsection{Implications}\label{implications}}

Our results showed that, for many controversial protest actions, there
is no consensus on what is and is not an acceptable means to advance a
cause. Instead, this judgement seems to depend on what the cause of a
protest is and how it aligns with one's beliefs and values. In this way,
our results confirm our contention that the distinction between
normative and non-normative forms of collective action should be
situated in the eye of the beholder. By developing an instrument for
measuring where people draw the line between acceptable and unacceptable
forms of collective action, we provided a paradigm that we hope will
stimulate research on when and why people differ in making this
distinction.

After all, this distinction can be consequential. If people believe that
even extreme actions are acceptable means to advance a cause they
support, especially if they moralize their support for that cause
(Mooijman et al., 2018), they might support violent extremism. In this
way, our findings could explain why 45\% of Republicans (2\% of
Democrats) supported the storming of the U.S. Capitol on January 6, 2021
(YouGov, 2021) even though it was illegal, destructive, and violent. If
people believe that even non-violent actions are unacceptable means to
advance a cause they oppose, they might support stifling dissent. In
this way, our findings could explain why Republicans responded to Black
Lives Matter protests by seeking to further criminalize disruptive
protest (e.g., blocking highways, Quinton, 2021) that might be most
effective at challenging injustice (Shuman et al., 2022, 2021). For
those reasons, researchers should further investigate when and why
people apply different standards when judging collective action for
various causes.

\hypertarget{limitations}{%
\subsection{Limitations}\label{limitations}}

Our research applied item response theory to judgements about collective
action to investigate whether where people draw the line between
acceptable and unacceptable means of protest depends on who the
protesters are and what they are protesting. In so doing, we moved the
distinction between normative and non-normative collective action from
the realm of the researcher's intuition into the realm of scientific
investigation, allowing us to test for double standards in judging
collective action.

Despite its strengths, our research is qualified by several limitations.
First, our research examined the hypothesized relationships for only two
political causes---protecting workers' rights, defunding the
police---and thus provided only limited evidence that our findings
generalize beyond those causes. Relying on few stimuli to establish an
effect threatens the validity and replicability of research findings and
severely constrains psychological research generalize beyond those few
stimuli (Yarkoni, 2022). Future research should sample a wider range of
causes to address this pervasive but often ignored problem (Judd et al.,
2012). This is particularly important when studying collective action as
recent theorizing (Jost et al., 2017) and research (Osborne et al.,
2019) highlighted the importance of differentiating between progressive
and conservative causes. Second, our research was based on two Western,
Educated, Industrialized, Rich, and Democratic (WEIRD; Henrich et al.,
2010) samples, which severely limits the cross-cultural generalizability
of our findings. Future research should include samples from a wider
range of cultural context to examine how the observed relationships vary
across cultural contexts.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Opinion polls show deep divides not only about the ends of protests, but
also about the means of protest. For example, only 11\% of Republicans
and 29\% of White Americans---compared to 59\% of Democrats and 66\% of
Black Americans---considered kneeling during the national anthem---a
non-violent form of collective action---an appropriate means to protest
racist police violence (YouGov, 2017). Our research showed that, in
contrast to well-documented double standards in other domains, people do
not apply a different standard when judging collective action by ingroup
members or for a cause aligned with the ingroup's interests. Instead,
our research demonstrated that people consider the same protest actions
to be more acceptable when the cause of the protest aligns with their
own ideological positions---in other words, that there are partisan
double standards in judging collective action.

@@

\leavevmode\vadjust pre{\hypertarget{ref-judd_treating_2012}{}}%
Judd, C. M., Westfall, J., \& Kenny, D. A. (2012). Treating stimuli as a
random factor in social psychology: {A} new and comprehensive solution
to a pervasive but largely ignored problem. \emph{Journal of Personality
and Social Psychology}, \emph{103}(1), 54--69.
\url{https://doi.org/10.1037/a0028347}

@@

\leavevmode\vadjust pre{\hypertarget{ref-mooijman_moralization_2018}{}}%
Mooijman, M., Hoover, J., Lin, Y., Ji, H., \& Dehghani, M. (2018).
Moralization in social networks and the emergence of violence during
protests. \emph{Nature Human Behaviour}, \emph{2}(6), 389--396.
\url{https://doi.org/10.1038/s41562-018-0353-0}
